Radio Crickets: Chirping Jets from Black Hole Binaries Entering their Gravitational Wave Inspiral

### 2015-12-02 Paper/paper.tex
Update documentation
Change figure filename
Resubmitted version

General relativisitic (GR) numerical simulations can predict the
binary's evolution in its final stage up to coalescence
\citep{2005PhRvL..95l1101P, 2006PhRvL..96k1102B, 2006PhRvL..96k1101C}.
This has stimulated interest in predicting related electromagnetic
signals of black hole coalescence.  Effects considered in the
literature include afterglows due to infall of gas on the remnant
black hole \citep{2005ApJ...622L..93M}, electromagnetic variability in
the circumbinary disk due to shocks induced by the sudden mass loss of
the binary following GW emission at coalescence
\citep{2007APS..APR.S1010B}, flares from shocked remnants of accretion
disk around the recoiled remnant black hole
\citep{2007PhRvL..99d1103L, 2008ApJ...682..758S, 2009CQGra..26i4032H},
quasi-periodic variability in gas accretion in the early stages of
black hole merger \citep{2008ApJ...672...83M, 2009CQGra..26i4032H,
  2012MNRAS.427.2680K}, evolution in the profiles of broad emission
lines shortly before merger \citep{2013MNRAS.432.1468M},
characteristic evolution in thermal emission due to circumbinary
accretion \citep{2015MNRAS.446L..36F}, brightening of the circumbinary
accretion disk due to viscous dissipation of gravitational waves
\citep{2008PhRvL.101d1101K}, and prompt tidal disruption of ambient
stars by the recoiled remnant \citep{2011MNRAS.412...75S}.

@@

In this Letter we study the morphology of a jet emitted by one of the
binary members during the binary's final phase of evolution before
coalescence, i.e., when the binary's dynamics is dominated by GW
emission.  We show that in this case the jet has a peculiar helical
chirping morphology that can be used to infer the binary parameters.

@@

We study a BH binary formed following a gas-rich merger of two
galaxies as a result of the processes described above.  We consider
black holes with masses $M_1$ and $M_2$ in a circular orbit of radius
$a$ around each other.  Let $M=M_1+M_2$.  For simplicity, we consider
an equal-mass binary ($M_1=M_2$) on a circular orbit. The orbital
speed of each black hole is given by \citep{2010PhRvD..81d7503L}
\begin{equation}
  v_\mathrm{orb} = \frac{1}{2}\left(\frac{GM}{a}\right)^{1/2}\!\!\!\! = 5.8\times 10^3\, \mathrm{km~s}^{-1}M_8^{1/2}a_{16}^{-1/2},
  \label{eqn:vorb}
\end{equation}
where $a_{16}\equiv (a/10^{16}$ cm$)$ and $M_8\equiv (M/10^8$
M$_\odot)$ are the binary separation and mass,
respectively.\footnote{Note that $10^{16} \mathrm{cm}\approx 3.4\times
  10^{-3} \mathrm{pc}$.  The Schwarzchild radius of a $10^8$~M$_\odot$
  black hole is $9.6\times 10^{-6}$~pc, or $\sim 3\times 10^{13}$~cm.}
The orbital period is
\begin{equation}
  P = 2\pi\left(\frac{GM}{a^3}\right)^{-1/2}\!\!\!\! = 1.72\, \mathrm{yr}\, a_{16}^{3/2} M_8^{-1/2}.
\end{equation}

In a gas-rich galaxy merger, the BH binary loses its angular momentum
by torquing the surrounding disk through spiral arms and expelling gas
out of a region twice as large as the binary orbit
\citep{2005ApJ...622L..93M, 2007PASJ...59..427H, 2008ApJ...672...83M,
  2009MNRAS.393.1423C}.  The expelled gas carries a specific angular
momentum of $\sim va$.  Further angular momentum dissipation of the
binary can happen due to its interaction with fresh gas that enters
the hollowed out region as a result of tidal torques
\citep{2008ApJ...672...83M, 2012ApJ...755...51N, 2012MNRAS.427.2680K,
  2012A&A...545A.127R}.  Therefore, the coalescence time of the binary
is inversely proportional to the rate at which fresh gas re-enters the
binary region.  But only a fraction of gas that enters the central
hollow region accretes onto the black hole and fuels quasar activity.
Therefore, we express $\dot M$ in Eddington units, $\dot{\mathcal
  M}\equiv\dot M /\dot M_E$, where $\dot M_E$ is the accretion rate
required to power the limiting Eddington luminosity with a radiative
efficiency of 10\%, $\dot M_E=2.3 M_\odot\, \mathrm{yr}^{-1}M_8$.  We
therefore have \citep{2010PhRvD..81d7503L}
\begin{equation}
  \frac{a}{\dot a} = t_\mathrm{gas}\approx\frac{J}{\dot Mva}=1.1\times 10^7\,
  \mathrm{yr}\,\dot{\mathcal M}^{-1},
\end{equation}
where $J=\mu va$ is the binary's angular momentum and $\mu=M/2$ is the
binary's reduced mass.  This time scale is greater by two orders of
magnitude than the time scale for coalescence due to angular momentum
loss by GW emission, which for circular orbits is given by
\citep{1964PhRv..136.1224P}
\begin{equation}
\frac{a}{\dot a} = t_\mathrm{GW} = \frac{5}{256}\frac{c^5a^4}{G^3M^2\mu}=2.53\times 10^5\,\mathrm{yr}\,\frac{a_{16}^4}{M_8^3}.
\end{equation}
By setting $t_\mathrm{gas}=t_\mathrm{GW}$, we can get the orbital
parameters for the binary when GWs take over.  The orbital speed at
this stage of the binary's evolution is given by
\begin{equation}
  v_\mathrm{orb} = 3.6\times 10^3\, \mathrm{km~s}^{-1}(\dot{\mathcal{M}}M_8)^{1/8},
  \label{eqn:v}
\end{equation}
The separation between the two black holes at this instant is given by
\begin{equation}
  a = 8.3\times 10^{-3}\, \mathrm{pc}\, M_8^{3/4}\dot{\mathcal{M}}^{-1/4}.
  \label{eqn:a}
\end{equation}
Note that the separation is dependent on the gas accretion rate.

\begin{figure}
  \begin{center}
    \begin{tikzpicture}
      \draw [-angle 45,dashed,thick,dash pattern = on 7pt off 2pt] (0,0) -- (3.0,0);
      \draw [-angle 45,dashed,thick,dash pattern = on 7pt off 2pt] (0,0) -- (0,4.0);
      \draw [-angle 45,thick] (0,0) -- (2.0,4.0);
      \draw [-angle 45,thick,red] (0,0) -- (5.0,4.0);
      \draw (1.0,2.0) arc (63:90:2.2);
      \draw (2.0,1.6) arc (38.6:63:2.56);
      \draw (0.5,2.5) node {\large $\chi$};
      \draw (1.90,2.3) node {\large $\psi$};
      \draw [dashed,thick,dash pattern = on 7pt off 2pt] (0,0) -- (3.0,6.0);
      \draw (0.0,4.5) node {\large $\mathbf{L}_\mathrm{orb}$};
      \draw (3.5,0.0) node {\large $\mathbf{v}_\mathrm{orb}$};
      \draw (1.0,3.0) node {\large $\mathbf{v}_\mathrm{jet}$};
      \draw (5.3,4.2) node {\large \textcolor{red}{$\mathbf{\tilde v}_\mathrm{jet}$}};
    \end{tikzpicture}
  \end{center}
  \caption{The orbital velocity $\mathbf{v}_\mathrm{orbital}$
    introduces the jet precision angle $\psi$.  Without the orbital
    motion, $\psi=0$, and the jet traces a cylindrical surface.  With
    the orbital motion, the net red velocity vector oscillates around
    the solid black vector and produces a conical surface with $\psi$
    as its half-opening angle.  Note that $v_\mathrm{orb}$ is
    exagerated here.  In practice, $v_\mathrm{orb}\ll
    v_\mathrm{jet}\sim c$.  The figure indicates that
    $\sin{\psi}\sim\tan{\psi}=v_\mathrm{orb}\cos{\chi}/v_\mathrm{jet}$.}
  \label{fig:jet}
\end{figure}

In the GW dominated phase of binary's evolution, recent simulations
show that the BHs can continue to accrete gas from the circumbinary
disk all the way until merger \citep{2015MNRAS.446L..36F}.  At this
stage, we consider a ballistic jet emanating from one of binary's
member BHs at speed $\beta=v_\mathrm{jet}/c$ at an angle $\chi$ to the
direction of the orbital angular momentum.  This geometry is depicted
in Figures~\ref{fig:BH_orbit} and \ref{fig:jet}.  As a result of the
binary orbit, the jet will wiggle on a conical surface.  (In the limit
of a small binary velocity, i.e., at large relative separations, the
jet material will move on a cylindrical surface.)  The half-opening
angle of the cone is given by \citep{1993ApJ...409..130R}
\begin{equation}
  \sin\psi = \frac{v_\mathrm{orb}\cos\chi}{v_\mathrm{jet}}.
\end{equation}

\begin{figure}
  \begin{center}
    \includegraphics*[width=\columnwidth]{evolve.pdf}
    % gw.py (case 5) 
  \end{center}
  \caption{Evolution of the binary parameters.  As the binary
    separation $a$ decreases, the orbital velocity $v_\mathrm{orb}$
    and the opening angle $\psi$ of the precessing jet, increase.
    This example describes the evolution of a binary with
    $M=10^8$~M$_\odot$ and $a_0=8.3\times 10^{-3}$~pc, but can be
    scaled to other masses and initial separations.}
  \label{fig:evolve}
\end{figure}

When the binary separation reaches the point where gravitational
wave-induced inspiral begins, the binary has a velocity given by
equation~(\ref{eqn:v}), and a separation given by
equation~(\ref{eqn:a}).  From equation (\ref{eqn:vorb}), the orbital
velocity of each BH evolves as
\begin{equation}
  \frac{\dot v_\mathrm{orb}}{v_\mathrm{orb}} = -\frac{1}{2}\frac{\dot a}{a}.
\end{equation}
which gives the evolution of the opening angle of the cone as 
\begin{equation}
 \dot\psi = -\frac{1}{2\cot\psi}\frac{\dot a}{a}.
\end{equation}
This can be solved analytically to get
\begin{equation}
  \sin\psi = \sin\psi_0\left(\frac{a_0}{a}\right)^{1/2},
\end{equation}
where $\psi_0$ and $a_0$ are the initial values. Thus, as the binary
continues to shrink the BH orbital velocity grows and the opening
angle of the conical jet increases (cone opens wider).  The rate at
which this happens reflects the binary's dynamical evolution.
Figure~\ref{fig:evolve} shows the evolution of the binary separation,
orbital velocity, and the half-opening angle for a binary in the
GW-dominated phase.

@@

\begin{figure*}
\begin{center}
  \begin{tabular}{cc}
    % jet3d_check2.py (case 0) 
    \includegraphics*[scale=0.45]{{jet_i40_beta0.90_mdot1.00_full}.pdf} &
    % jet3d_image_fulljet.py (case 1) 
    \includegraphics*[scale=0.45]{jet_check_zoom1.pdf}
  \end{tabular}
\end{center}
\caption{The left panel shows forward and backward jets for an
  equal-mass binary with total mass $M=10^{10}$~M$_\odot$.  We have
  assumed $i=40^\circ$, $\beta=0.9$, and angular diameter distance $d
  = 100$~Mpc.  The jets are twisted because of the apparant
  superluminal motion.  Also evident is the opening of the jet and the
  radio jet chirp.  Right panel shows normalised brightness of the
  inner 30~mas region of the forward jet.}
\label{fig:jet_mdot1.0}
\end{figure*}

@@

Finally, we assume that the emission from each jet is optically thin
at the frequency of observation with a spectrum in the rest frame of
the particle given by $P(\nu)\propto \nu^{-\alpha}$
\citep{1982ApJ...262..478G}.  The observed flux density $S(\nu)$ in
erg cm$^{-2}$ Hz$^{-1}$ is given by
\begin{equation}
  S(\nu)=S_r(\nu)D^{3+\alpha},
\end{equation}
where $D$ is the Doppler shift parameter
$[\gamma(1-\beta\cos\phi)]^{-1}$ and $S_r$ is the rest frame flux
density.  Further, to consider the effects of jet evolution we assume
that $S_r$ decreases with time according to a simple power law
\begin{equation}
  S_r\propto t^{-\delta}.
\end{equation}
We assume $\delta=1$ and $\alpha=1$, hereafter
\citep{1982ApJ...262..478G}.

\begin{table}
\begin{center}
\begin{tabular}{c l}
\hline 
Parameter & Description \\ 
\hline 
$M$ & Total mass \\
$\zeta$ & Mass asymmetry \\
$\chi$ & Intrinsic jet direction \\
$e$ & Orbital eccentricity \\
$\alpha$ & SED slope \\
$\delta$ & Time evolution of brightness evolution \\
$\beta$ & Jet speed \\
$i$ & Orbital inclination \\ 
$\dot{\mathcal M}$ & Gas accretion rate\\
\hline
\end{tabular}
\end{center}
\caption{Parameters of the problem.}
\label{table:params}
\end{table}

\section{Flagging GW binaries and probing GR}

@@

The increasing orbital speed of the binary leads to an increase in the
jet's precession angular velocity.  This increases the helicity of the
jet such that the jet is wound up increasingly tightly closer to the
source.  This gives rise to a ``chirping'' morphology as seen in
Figure~\ref{fig:jet_mdot1.0}.  The chirp is easier to resolve in jets
that are closer to the line of sight, where superluminal motion can
magnify the forward jet.  A chirping black hole is an unambiguous
signature of a rapidly evolving binary BH system; the chirp is direct
reflection of the orbital velocity evolution in
Figure~\ref{fig:evolve}.  Together, the three features of the jet
morphology in Figure~\ref{fig:jet_mdot1.0}, namely, conical jet
geometry with an increasing opening angle, a radio chirp geometry, and
an asymmetrical shape, directly probe the dynamics of the binary in
its GW phase.

Observations of such a milliarcsecond geometry can thus constrain GR
in the strong field limit.  The relative separation of the jet winding
directly constraints the ratio $a_{16}^4/M_8^3$ by constraining the
evolution of the orbital velocity.  The jet opening angle evolution
tracks the evolution of the binary separation.  We make the code for
implementing the model presented in this Letter available at
\url{https://github.com/gkulkarni/JetMorphology} under the MIT open
source software license.  This code can be used to explore the
dynamical, astrophysical, and geometrical parameter spaces of the
model.

>>>>

General relativisitic (GR) numerical simulations can predict the
binary's evolution in its final stage up to coalescence
\citep{2005PhRvL..95l1101P, 2006PhRvL..96k1102B, 2006PhRvL..96k1101C}.
This has stimulated interest in predicting related electromagnetic
signals of black hole coalescence.  Effects considered in the
literature include afterglows due to infall of gas on the remnant
black hole \citep{2005ApJ...622L..93M}, electromagnetic variability in
the circumbinary disk due to shocks induced by the sudden mass loss of
the binary following GW emission at coalescence
\citep{2007APS..APR.S1010B, 2008ApJ...676L...5L}, flares from shocked
remnants of accretion disk around the recoiled remnant black hole
\citep{2007PhRvL..99d1103L, 2008ApJ...682..758S, 2009CQGra..26i4032H},
quasi-periodic variability in gas accretion in the early stages of
black hole merger \citep{2008ApJ...672...83M, 2009CQGra..26i4032H,
  2012MNRAS.427.2680K, 2013MNRAS.436.2997D, 2014ApJ...783..134F,
  2015ApJ...807..131S}, evolution in the profiles of broad emission
lines shortly before merger \citep{2013MNRAS.432.1468M},
characteristic evolution in thermal emission due to circumbinary
accretion \citep{2015MNRAS.446L..36F}, brightening of the circumbinary
accretion disk due to viscous dissipation of gravitational waves
\citep{2008PhRvL.101d1101K}, and prompt tidal disruption of ambient
stars by the recoiled remnant \citep{2011MNRAS.412...75S}.

@@

In this Letter, we study the morphology of jets from binary BHs whose
dynamics is doiminated by GW emission.  Jets in binary BHs have been
studied in a number of previous works \citep{1980ApJ...238L.129S,
  1981ApJ...246L..65I, 1981ApJ...246L.141H, 1982ApJ...258L..63G,
  1982ApJ...253L...1G, 1982ApJ...262..478G}.  It has been shown that
the orbital motion of binary results in a precession of the jet,
resulting in a helical or ``wiggly'' jet morphology.  Such models of
jet precession have been used to understand the observed jet
morphology in several AGN.  For example, recently
\citet{2014MNRAS.445.1370K} fit a jet precession model to the observed
jet in the quasar S5 1928$+$738 using 18 yr of very long baseline
interferometric (VLBI) data.  This object was first studied using BH
binary models by \citet{1993ApJ...409..130R}, but
\citet{2014MNRAS.445.1370K} also included effects of BH spin.  Other
objects studied in this manner are PG 1302-102
\citep{2015MNRAS.454.1290K, 2015Natur.518...74G}, BL Lacertae
\citep{2003MNRAS.341..405S, 2013MNRAS.428..280C}, NGC 4151
\citep{2012ApJ...759..118B}, OJ 287 \citep{2012MNRAS.421.1861V}, S5
1803$+$784 \citep{2008A&A...483..125R}, 3C 345
\citep{2005A&A...431..831L}, 3C 210 \citep{2004MNRAS.349.1218C}, PKS
420-014 \citep{2001A&A...374..784B}, 3C 273
\citep{2000A&A...360...57R}, and Mrk 501 \citep{1999A&A...347...30V}.
\citet{1982ApJ...262..478G} also studied several AGN using their model
of jet precession.

The shortest binary separations inferred from these studies are of the
order of $10^{-2}$ pc.  At smaller separations of around $10^{-3}$ pc,
the orbital velocity of the BHs typically becomes relativistic
($v_\mathrm{orb}\sim 10^4$ km$/$s or $\beta\sim 0.01$).  It was
thought that at these small separations, the binary evolves in a
vacuum, because the inspiral time scale of the binary is likely
shorter than the viscous time scale of any circumbinary gas disk.  As
a result, gas fails to catch up with the BHs, making any accretion
activity impossible in this stage of the binary's evolution
\citep{2005ApJ...622L..93M, 2010PhRvD..81b4019S, 2010ApJ...714..404T}.
However recent viscous hydrodynamical simulations do not find this
decoupling between the BHs and the circumbinary disk.  In these
simulations, the binary's member BHs can continue to accrete gas from
the circumbinary accretion disk all the way until coalescence
\citep{2012PhRvL.109v1102F, 2013MNRAS.436.2997D, 2014PhRvD..89f4060G,
  2014PhRvD..90j4030G, 2015MNRAS.446L..36F, 2015ApJ...807..131S}.
This opens up the possibility of jet production by one of the binary
members during the binary's final phase of evolution before
coalescence, i.e., when the binary's dynamics is dominated by GW
emission.  In this Letter we study the morphology of such jets.

All of the above models of jet precession in binary BHs consider
stable binaries, in which the time scale of evolution of the binary
separation is much larger than the orbital period of the binary.  For
such binaries, the helical morphology of the jet is does not change
significantly between precession cycles.  However, we will see below
that for a binary BH with separation $a$, the orbital period is
proportional to $a^{1.5}$, while the time scale of evolution of the
binary separation is proportional to $a^{4}$.  Thus, as the binary
shrinks, the latter time scale decreases much more rapidly than the
former, and for small enough separations, the two time scales can be
comparable.  In this paper, we study jets in this previously ignored
regime.  We show that at this stage of the binary's evolution, the jet
morphology is more complicated: the jet now has a peculiar helical
chirping morphology.  We further argue that in this case the jet
morphology can be used to infer the binary parameters and show how
such features can be distinguished from the usual helical jet
morphology.

@@

We study a BH binary formed following a gas-rich merger of two
galaxies as a result of the processes described above.  We consider
black holes with masses $M_1$ and $M_2$ in a circular orbit of radius
$a$ around each other.  Let $M=M_1+M_2$.  For simplicity, we consider
an equal-mass binary ($M_1=M_2$) on a circular orbit. The orbital
speed of each black hole is given by 
\begin{equation}
  v_\mathrm{orb} = \frac{1}{2}\left(\frac{GM}{a}\right)^{1/2}\!\!\!\! = 5.8\times 10^3\, \mathrm{km~s}^{-1}M_8^{1/2}a_{16}^{-1/2},
  \label{eqn:vorb}
\end{equation}
where $a_{16}\equiv (a/10^{16}$ cm$)$ and $M_8\equiv (M/10^8$
M$_\odot)$ are the binary separation and mass,
respectively.\footnote{Note that $10^{16} \mathrm{cm}\approx 3.4\times
  10^{-3} \mathrm{pc}$.  The Schwarzchild radius of a $10^8$~M$_\odot$
  black hole is $9.6\times 10^{-6}$~pc, or $\sim 3\times 10^{13}$~cm.}
The orbital period is
\begin{equation}
  P = 2\pi\left(\frac{GM}{a^3}\right)^{-1/2}\!\!\!\! = 1.72\, \mathrm{yr}\, a_{16}^{3/2} M_8^{-1/2}.
  \label{eqn:period}
\end{equation}

In a gas-rich galaxy merger, the BH binary loses its angular momentum
by torquing the surrounding disk through spiral arms and expelling gas
out of a region twice as large as the binary orbit
\citep{1994ApJ...421..651A, 2005ApJ...622L..93M, 2007PASJ...59..427H,
  2008ApJ...672...83M, 2009MNRAS.393.1423C}.  The expelled gas carries
a specific angular momentum of $\sim va$.  Further angular momentum
dissipation of the binary can happen due to its interaction with fresh
gas that enters the hollowed out region as a result of tidal torques
\citep{2008ApJ...672...83M, 2012ApJ...755...51N, 2012MNRAS.427.2680K,
  2012A&A...545A.127R}.  Therefore, the coalescence time of the binary
is inversely proportional to the rate at which fresh gas re-enters the
binary region.  But only a fraction of gas that enters the central
hollow region accretes onto the black hole and fuels quasar activity.
Therefore, we express $\dot M$ in Eddington units, $\dot{\mathcal
  M}\equiv\dot M /\dot M_E$, where $\dot M_E$ is the accretion rate
required to power the limiting Eddington luminosity with a radiative
efficiency of 10\%, $\dot M_E=2.3 M_\odot\, \mathrm{yr}^{-1}M_8$.  We
therefore have \citep{2010PhRvD..81d7503L, 2015MNRAS.448.3603D}
\begin{equation}
  \frac{a}{\dot a} = t_\mathrm{gas}\approx\frac{J}{\dot Mva}=1.1\times 10^7\,
  \mathrm{yr}\,\dot{\mathcal M}^{-1},
\end{equation}
where $J=\mu va$ is the binary's angular momentum and $\mu=M/2$ is the
binary's reduced mass.  This time scale is greater by two orders of
magnitude than the time scale for coalescence due to angular momentum
loss by GW emission, which for circular orbits is given by
\citep{1964PhRv..136.1224P}
\begin{equation}
  \frac{a}{\dot a} = t_\mathrm{GW} = \frac{5}{256}\frac{c^5a^4}{G^3M^2\mu}=2.53\times 10^5\,\mathrm{yr}\,\frac{a_{16}^4}{M_8^3}.
  \label{eqn:shrink}
\end{equation}
By setting $t_\mathrm{gas}=t_\mathrm{GW}$, we can get the orbital
parameters for the binary when GWs take over.  The orbital speed at
this stage of the binary's evolution is given by
\begin{equation}
  v_\mathrm{orb} = 3.6\times 10^3\, \mathrm{km~s}^{-1}(\dot{\mathcal{M}}M_8)^{1/8},
  \label{eqn:v}
\end{equation}
The separation between the two black holes at this instant is given by
\begin{equation}
  a = 8.3\times 10^{-3}\, \mathrm{pc}\, M_8^{3/4}\dot{\mathcal{M}}^{-1/4}.
  \label{eqn:a}
\end{equation}
Note that the separation is dependent on the gas accretion rate.

\begin{figure}
  \begin{center}
    \begin{tikzpicture}
      \draw [-angle 45,dashed,thick,dash pattern = on 7pt off 2pt] (0,0) -- (3.0,0);
      \draw [-angle 45,dashed,thick,dash pattern = on 7pt off 2pt] (0,0) -- (0,4.0);
      \draw [-angle 45,thick] (0,0) -- (2.0,4.0);
      \draw [-angle 45,thick,red] (0,0) -- (5.0,4.0);
      \draw (1.0,2.0) arc (63:90:2.2);
      \draw (2.0,1.6) arc (38.6:63:2.56);
      \draw (0.5,2.5) node {\large $\chi$};
      \draw (1.90,2.3) node {\large $\psi$};
      \draw [dashed,thick,dash pattern = on 7pt off 2pt] (0,0) -- (3.0,6.0);
      \draw (0.0,4.5) node {\large $\mathbf{L}_\mathrm{orb}$};
      \draw (3.5,0.0) node {\large $\mathbf{v}_\mathrm{orb}$};
      \draw (1.0,3.0) node {\large $\mathbf{v}_\mathrm{jet}$};
      \draw (5.3,4.2) node {\large \textcolor{red}{$\mathbf{\tilde v}_\mathrm{jet}$}};
    \end{tikzpicture}
  \end{center}
  \caption{The orbital velocity $\mathbf{v}_\mathrm{orb}$
    introduces the jet precession angle $\psi$.  Without the orbital
    motion, $\psi=0$, and the jet traces a cylindrical surface.  With
    the orbital motion, the net red velocity vector oscillates around
    the solid black vector and produces a conical surface with $\psi$
    as its half-opening angle.  Note that $v_\mathrm{orb}$ is
    exagerated here.  In practice, $v_\mathrm{orb}\ll
    v_\mathrm{jet}\sim c$.  The figure indicates that
    $\sin{\psi}\sim\tan{\psi}=v_\mathrm{orb}\cos{\chi}/v_\mathrm{jet}$.}
  \label{fig:jet}
\end{figure}

In the GW dominated phase of binary's evolution, recent simulations
show that the BHs can continue to accrete gas from the circumbinary
disk all the way until merger \citep{2015MNRAS.446L..36F}.  At this
stage, we consider a ballistic jet emanating from one of binary's
member BHs at speed $\beta=v_\mathrm{jet}/c$ at an angle $\chi$ to the
direction of the orbital angular momentum.  This geometry is depicted
in Figures~\ref{fig:BH_orbit} and \ref{fig:jet}.  As a result of the
binary orbit, the jet will wiggle on a conical surface.  (In the limit
of a small binary velocity, i.e., at large relative separations, the
jet material will move on a cylindrical surface.)  The half-opening
angle of the cone is given by \citep{1993ApJ...409..130R}
\begin{equation}
  \sin\psi = \frac{v_\mathrm{orb}\cos\chi}{v_\mathrm{jet}}.
\end{equation}
(Note that this relation is valid in the $v_\mathrm{orb}\ll
v_\mathrm{jet}\sim c$. Also, we are ignoring any relativistic
corrections to this equation.)

When the binary separation reaches the point where gravitational
wave-induced inspiral begins, the binary has a velocity given by
equation~(\ref{eqn:v}), and a separation given by
equation~(\ref{eqn:a}).  From equation (\ref{eqn:vorb}), the orbital
velocity of each BH evolves as
\begin{equation}
  \frac{\dot v_\mathrm{orb}}{v_\mathrm{orb}} = -\frac{1}{2}\frac{\dot a}{a}.
\end{equation}
which gives the evolution of the opening angle of the cone as 
\begin{equation}
  \dot\psi = -\frac{1}{2\cot\psi}\frac{\dot a}{a}.
  \label{eqn:widen}
\end{equation}
This can be solved analytically to get
\begin{equation}
  \sin\psi = \sin\psi_0\left(\frac{a_0}{a}\right)^{1/2},
\end{equation}
where $\psi_0$ and $a_0$ are the initial values. Thus, as the binary
continues to shrink the BH orbital velocity grows and the opening
angle of the conical jet increases (cone opens wider).  The rate at
which this happens reflects the binary's dynamical evolution.

@@

\begin{figure*}
\begin{center}
  \begin{tabular}{cc}
    % jet.py (case 0) 
    \includegraphics*[scale=0.45]{{jet_i40_beta0.90_mdot1.00_full}.pdf} &
    % jet_image.py (case 1) 
    \includegraphics*[scale=0.45]{{jet_i40_beta0.90_mdot1.00_image_zoom1}.pdf}
  \end{tabular}
\end{center}
\caption{The left panel shows forward and backward jets for an
  equal-mass binary with total mass $M=10^{10}$~M$_\odot$.  (This
  figure shows the position of many jet particles observed at a fixed
  time on Earth.)  We have assumed $i=40^\circ$, $\beta=0.9$, and
  angular diameter distance $d = 100$~Mpc.  The jets are twisted
  because of the apparant superluminal motion.  Also evident is the
  opening of the jet and the radio jet chirp.  Right panel shows
  normalised brightness of the inner 30~mas region of the forward
  jet. (This panel shows the intensity of many particles observed at a
  fixed time on Earth.)}
\label{fig:jet_mdot1.0}
\end{figure*}

@@

Finally, we assume that the emission from each jet is optically thin
at the frequency of observation with a spectrum in the rest frame of
the particle given by $P(\nu)\propto \nu^{-\alpha}$
\citep{1982ApJ...262..478G}.  The observed flux density $S(\nu)$ in
erg cm$^{-2}$ Hz$^{-1}$ is given by
\begin{equation}
  S(\nu)=S_r(\nu)D^{3+\alpha},
\end{equation}
where $D$ is the Doppler shift parameter
$[\gamma(1-\beta\cos\phi)]^{-1}$ and $S_r$ is the rest frame flux
density.  Further, to consider the effects of jet evolution we assume
that $S_r$ decreases with time according to a simple power law
\begin{equation}
  S_r\propto t_\mathrm{proper}^{-\delta},
\end{equation}
where $t_\mathrm{proper}$ is the proper time in the frame of the jet.
We assume $\delta=1$ and $\alpha=1$, hereafter
\citep{1982ApJ...262..478G}.

\section{Jet Morphology}

@@

The increasing orbital speed of the binary leads to an increase in the
jet's precession angular velocity.  This increases the helicity of the
jet such that the jet is wound up increasingly tightly closer to the
source.  This gives rise to a ``chirping'' morphology as seen in
Figure~\ref{fig:jet_mdot1.0}.  The chirp is easier to resolve in jets
that are closer to the line of sight, where superluminal motion can
magnify the forward jet.  A chirping black hole is an unambiguous
signature of a rapidly evolving binary BH system; the chirp is direct
reflection of the orbital velocity evolution of the binary. Together,
the three features of the jet morphology in
Figure~\ref{fig:jet_mdot1.0} are:
\begin{itemize}
\item Increase in the helicity of the jet towards the central source (``chirpyness''),
\item Widening of the opening angle towards the central source, and 
\item An apparant variation in the jet precession axis towards the central source (``twist''). 
\end{itemize}
These directly probe the dynamics of the binary in its GW phase.

\subsection{Can chirping jets be observed?}

Observations of such a milliarc second scale geometry with the three
features described above can thus flag a BH binary nearing the end of
its merger.  We now consider the question of whether such morphologies
will be measurable in real observations.

As discussed in Section~\ref{sec:intro}, several helical jets have
been observed thanks to VLBI.  A resolution of few tens of microarc
seconds can be achieved using the Very Long Baseline Array (VLBA) with
sensitivity of up to 30 $\mu$Jy$/$beam, as demonstrated by the MOJAVE
survey \citep{2009AJ....138.1874L}.  VLA has angular resolution of few
tens of milliarcsecconds with a sensitivity of about 100
$\mu$Jy$/$beam.  The European VLBI network (EVN) which can resolve up
to 0.13 milliarc seconds and can detect surface brightness down to 19
$\mu$Jy$/$beam.  The Very Large Array (VLA) has an angular resolution
of few tens of milliarcsecconds with a sensitivity of about 100
$\mu$Jy$/$beam.  The future Square Kilometer Array (SKA) on its own
will have 2 milliarc second resolution at around 10 GHz, but SKA will
be an order of magnitude more sensitive than current instruments with
a sensitivity of about 50 nJy$/$beam.  Thus angular resolution of tens
to hundreds of microarc seconds and flux sensitivity of a few
$\mu$Jy$/$beam is achievable with current VLBI instruments.

The brighter part of the jet (intensity within factor of 10 from the
maximum) in Figure~\ref{fig:jet_mdot1.0} (right panel) has separations
of $\sim 1$ mas.  It should therefore be possible to resolve the jet
at an angular resolution of $\sim 2$ mas.  (In the next section, we
discuss an algorithm to fit the jet model to observational data.)  If
it is aligned closer to the line of sight, the forward jet is
magnified due to apparant superluminal motion.  This will make it
possible to resolve jets of objects that are farther away than the
example in Figure~\ref{fig:jet_mdot1.0}.  Binaries with higher masses
have large orbital periods that will result is a helix with wider
seperation between its cycles.  These objects would therefore be
easier to detect.  We have not modelled the absolute surface
brightness of jets, only relative variation in the jet brightness (as
shown in Figure~\ref{fig:jet_mdot1.0}), but due to the $(1+z)^{-4}$
dimming, it will be exceedingly difficult to detect such jets at
higher redshifts.

We note here that there is at least one object in the literature that
shows a jet morphology similar to Figure~\ref{fig:jet_mdot1.0}.  The
quasar 4C 18.68 described by \citet{1982ApJ...253L...1G} shows a
helical jet with an apparant opening angle that increases towards the
central source from 11$^\circ$ to 45$^\circ$ in just two cycles
\citet{1982ApJ...262..478G}.  The observing beamwidth of about 400 mas
makes it difficult to resolve a chirp in this source, but it is a
prototype of the morphology that we are discussing here.
\citet{2015MNRAS.453.1562G} have reported that 111 candidates show
periodic variability in brightness in the Cataline Real-time Transient
Survey, indicating presence of binary BHs.  An analysis of these
candidates suggests the presence of circumbinary gas at small orbital
distances \citep{2015MNRAS.453.1562G}.  This bodes well for the
likelihood of the presence of chirpy jets.  In low resolution wide
field survey, detecting modulations in the lightcurves would be an
easier diagnostic of a binary BH \citep{2011ApJ...743..136O,
  2011ApJ...734L..37K}.

\subsection{Deriving binary parameters from observations of chirping jets}

From observations of chirping jets, we can infer the dynamical
parameters of the binary BH by first fitting a purely geometric model
to the observed jet morphology.  To create a model of this kind, we
generalise the simple helical model considered by
\citet{2014MNRAS.445.1370K} and \citet{2001NewA....6...61N} to write
\begin{align}
  z &= x^{-\alpha} \frac{B}{2\pi} u \cos{u}, \\
  y &= x^{-\alpha} \frac{B}{2\pi} u \sin{u}, \\
  x &= \frac{A}{2\pi} u \left(\frac{u}{2\pi}\right)^\beta.
\end{align}
Here $u$ is the azimuthal angle around the jet axis, $x$ and $y$ axes
are in the plane of the sky, and the $z$ axis points to the observer.
The $x^{-\alpha}$ term in the first two equations captures the
widening opening angle of the jet, while the $(u/2\pi)^{\beta}$ term
in the last equation parameterises the chirp.  This model can be fit
to an observed jet on the sky by using two rotation matrices,
$\mathcal{R}_\lambda$ and $\mathcal{R}_i$, which rotate the model
in the above equtions to match with the jet precession axis and the
jet symmetry axis.  These matrices are given by
